---
title: "Pressures to biodiversity across South African biomes as captured through land cover data."
title-block-banner: "#A7A9AC"
title-block-banner-color: white
author:
  - name:
      given: Kagiso K.
      family: Mogajane
    orcid: 0000-0003-4131-5634
    affiliations:
      - ref: af1
      
  - name:
      given: Anisha
      family: Dayaram
    orcid: 0000-0003-4131-5634
    affiliations:
      - ref: af1
      - ref: af2
      
  - name:
      given: Andrew L.
      family: Skowno
    orcid: 0000-0003-4131-5634
    affiliations:
      - ref: af1
      - ref: af3
      
  - name:
      given: Stephni
      family: van der Merwe
    orcid: 0000-0003-4131-5634
    affiliations:
      - ref: af1
      - ref: af4
      
  - name:
      given: Maphale
      family: Monyeki
    orcid: 0000-0003-4131-5634
    affiliations:
      - ref: af1

affiliations:
  - id: af1
    number: 1
    name: South African National Biodiversity Institute
  - id: af2
    number: 2
    name: Restoration and Conservation Biology Research Group, Centre for African Ecology, School of Animal, Plant & Environmental Sciences, University of the Witwatersrand, Johannesburg, South Africa
  - id: af3
    number: 3
    name: Plant Conservation Unit, Department of Biological Sciences, University of Cape Town, Cape Town, South Africa
  - id: af4
    number: 4
    name: Centre for Statistics in Ecology, Environment and Conservation, Department of Biological Sciences, University of Cape Town, Cape Town, South Africa 

date: last-modified

bibliography: ter_lcc.bib

citation:
  type: webpage
  container-title: National Biodiversity Assessment 2025
  publisher: South African National Biodiversity Institute
  url: http://nba.sanbi.org.za/

appendix-cite-as: false
google-scholar: true
---

------------------------------------------------------------------------

## Summary

Recent habitat loss from 1990 to 2022 varies greatly across biomes. Land cover data (Table 1) from 2022, shows that the Indian Ocean Coastal Belt had the highest habitat loss between with 21% (1 026 km2) loss, 8.5% (17 930 km2) of the Grassland, and 5.9% (3 453 km2) of the Fynbos, the Savanna 4.1% (13 549 km2). While all other biomes have experienced less than 4% habitat loss between 1990 and 2022, this still contributes to a steady, long-term reduction in natural habitat across the country. This reflects a persistent national trend of habitat decline, even in biomes where change may be subtle or difficult to detect.

```{r}

library(readxl)
library(dplyr)
library(gt)
library(writexl)
library(scales)  # for number formatting

# Load data
RNHL <- read_excel("C:/Users/K.Mogajane/OneDrive - sanbi.org.za/Kagiso Mogajane/Research Assistant 2025/LandCoverChangeAnalysis_NBA2025/R_Project/nba_landcover/data/Biome_Summaries/RemainingNaturalHabitat_and_Loss.xlsx")

# Rename columns with superscript ² and consistent formatting
colnames(RNHL) <- colnames(RNHL) %>%
  gsub("Extent \\(km2\\) 1750 \\(Reference\\)", "1750 extent (km²)", .) %>%
  gsub("Extent \\(km2\\) 1990", "1990 extent (km²)", .) %>%
  gsub("Extent \\(km2\\) 2014", "2014 extent (km²)", .) %>%
  gsub("Extent \\(km2\\) 2022", "2022 extent (km²)", .) %>%
  gsub("Recent Loss [Kk]m2 \\(1990[- ]to[- ]2022\\)", "Recent loss 1990 to 2022 (km²)", .) %>%
  gsub("Recent Loss Km2 \\(1990[- ]2022\\)", "Recent loss 1990 to 2022 (km²)", .) %>%
  gsub("Percentage Loss \\(1990[- ]2022\\)", "Percentage loss 1990 to 2022", .)

# Format percentage column
RNHL <- RNHL %>%
  mutate(
    `Percentage loss 1990 to 2022` = paste0(
      round(
        as.numeric(gsub("%", "", trimws(`Percentage loss 1990 to 2022`))),
        1
      ),
      "%"
    )
  )

# Optional: save formatted Excel
write_xlsx(RNHL, "RemainingNaturalHabitat_and_Loss_formatted.xlsx")

# GT theme function
theme_my_gt <- function(GT_TBL) {
  last_row <- nrow(GT_TBL[["_data"]])
  
  GT_TBL %>%
    gt::tab_style(
      style = list(
        gt::cell_fill(color = "lightgray"),
        gt::cell_borders(
          sides = c("top", "bottom"),
          color = "black",
          weight = gt::px(2),
          style = "solid"
        ),
        gt::cell_borders(
          sides = c("left", "right"),
          color = "lightgray"
        ),
        gt::cell_text(weight = "bold")
      ),
      locations = gt::cells_column_labels()
    ) %>%
    gt::tab_style(
      style = gt::cell_borders(
        sides = c("top", "bottom"),
        color = "gray50",
        weight = gt::px(1),
        style = "dotted"
      ),
      locations = gt::cells_body()
    ) %>%
    gt::tab_style(
      style = gt::cell_borders(
        sides = c("left", "right"),
        color = "transparent"
      ),
      locations = gt::cells_body()
    ) %>%
    gt::tab_style(
      style = gt::cell_borders(
        sides = "bottom",
        color = "black",
        weight = gt::px(2),
        style = "solid"
      ),
      locations = gt::cells_body(rows = last_row)
    )
}

# Create GT table with thousand separators and alignment
RNHL %>%
  gt() %>%
  tab_caption(
    md("**Table 1.** Natural habitat remaining and habitat loss by biome from 1750 to 2022, showing biome-level totals and changes between the years. The 1750 extent represents a pre-industrial reference baseline, and totals are based on the national land cover datasets for 1990 and 2022.")) %>%
  theme_my_gt() %>%
  gt::fmt_number(
    columns = c(
      `1750 extent (km²)`,
      `1990 extent (km²)`,
      `2014 extent (km²)`,
      `2022 extent (km²)`,
      `Recent loss 1990 to 2022 (km²)`
    ),
    decimals = 0,
    sep_mark = " "
  ) %>%
  gt::cols_align(
    align = "right",
    columns = `Percentage loss 1990 to 2022`
  )

```

## Biome summaries

::: {.panel-tabset group="language"}
### Albany Thicket

The Albany Thicket biome is a dense assemblage of evergreen succulent and woody shrubs and trees, mainly occuring in the south-east South Africa. It is an ancient biome believed to have been widespread in the Eocene @cowling2005 and forms part of the Maputaland-Pondoland-Albany Global Biodiversity Hotspot.

The biome has lost 1.9% of its natural habitat between 1990 and 2022 (Table 1). Historically, croplands were responsible for large losses in the extent of natural habitat compared to other pressures. Cropland decreased in extent as both secondary natural areas and build-up areas have increased (Figure 1).

Land cover data from 1990 to 2022 (Figure 1) indicate several distinct trends in the Albany Thicket biome. Between 1990 and 2014, cropland extent declined steadily, while secondary-natural vegetation increased (Figure 1), likely reflecting a shift away from crop production towards wildlife ranching and rangeland herbivory @lloyd2002. Built-up areas and artificial waterbodies also declined slightly during this period, while plantations and mining areas remained largely unchanged (Figure 1). However, from 2014 to 2022 this pattern shifted to croplands expanding, coinciding with a decline in secondary-natural vegetation particularly in areas associated with irrigation development and citrus farming @hoare2006 @carvalho2022. Built-up areas increased during this later period, while mining showed and plantation extents showed only minor increases.

The expansion of areas classified as secondary-natural reflects intersecting historical, political and ecological processes. Declines in crop production, including the contraction of pineapple farming in the mid-20th century, combined with shifts in land ownership and land-use priorities during the 1990s land reform period, led to portions of former croplands transitioning out of intensive production and gradually revegetating @stickler2015. In communal and peri-urban landscapes, revised management objectives on urban commonages further encouraged low-intensity use, such as subsistence wood harvesting, which indirectly supported vegetation regrowth. However, these same socio-economic transitions also facilitated increased goat farming and broader pastoral pressures, particularly within succulent Thicket vegetation, where sustained browsing degrades the natural structure and drives transitions toward an altered state that resemble pseudo-karoo or grassy states, thus captured as secondary-natural @mills2004 @lechmere-oertel2004.

Land cover data from 1990 to 2022 (Figure 1) show clear trends in the Albany Thicket biome. Between 1990 and 2014, cropland steadily declined while secondary-natural vegetation expanded, likely reflecting a shift away from crop production towards wildlife ranching and rangeland herbivory @lloyd2002. Built-up areas and artificial waterbodies decreased slightly, while plantations and mining remained largely unchanged. From 2014 to 2022, this pattern reversed: cropland expanded and secondary-natural vegetation declined, particularly in areas linked to irrigation development and citrus farming @hoare2006 @carvalho2022. Built-up areas increased during this period, while plantations and mining showed only minor growth.

The increase in secondary-natural vegetation is rooted in historical, political, and ecological processes. Declines in crop production, including the contraction of pineapple farming in the mid-20th century, combined with land reform in the 1990s, led to former croplands transitioning out of intensive production and gradually revegetating @stickler2015. In communal and peri-urban areas, revised management objectives on urban commonages encouraged low-intensity use, such as subsistence wood harvesting, which indirectly supported regrowth. However, these socio-economic shifts also drove increased goat farming and broader pastoral pressures, particularly in succulent Thicket vegetation, where sustained browsing degrades natural structure and pushes the system towards altered states resembling pseudo-karoo or grassy conditions, recorded as secondary-natural @mills2004 @lechmere-oertel2004.

There is evidence that shows that passive revegetation can facilitate the recovery of woody thicket vegetation on previously mined land, with topsoil transfers and natural regeneration producing substantial increases in woody cover and moderate similarity to intact thicket @hall2003, although gradually.

It is important to note that while land cover changes in recent decades have not been pronounced, most of the arid and valley thicket ecosystem types were cleared and overutilised before satellite monitoring started. Historical overstocking of livestock has likely led to irreversible vegetation changes in the thicket biome @lechmere-oertel2005 @rutherford2014. Because thicket, especially the succulent and slow-going components, has low regenerative capacity, the vegetation cannot return to its previously intact state without human intervention @lechmere-oertel2005 @carvalho2022. Luckily, restoration efforts can be successful when active restoration is applied with protection of the vegetation over time @mills2006 @del_rio-mena2021 @galuszynski2022. However, continued goat farming, of which South Africa is the leading producer of mohair, and sheep farming remains key pressures to the remaining rangelands in thicket. Therefore, what is currently mapped as “natural” and “secondary natural” in the land cover data sets, might reflect thicket in various stages of degradation, not captured by the land cover data sets. The 2022 land cover data (Figure 2) show that majority of the thicket is classified more natural than non-natural, however, within the non-natural classification, the key pressure in the biome is cropland, secondary-natural vegetation which might be in different degradation states @mills2004 @LeaverCherry2020 (Figure 2).

```{r}
# Load libraries
library(readxl)
library(dplyr)
library(ggplot2)

# Define file path
file_path <- "C:/Users/K.Mogajane/OneDrive - sanbi.org.za/Kagiso Mogajane/Research Assistant 2025/LandCoverChangeAnalysis_NBA2025/R_Project/nba_landcover/data/Biome_Summaries/Albany_Thicket.xlsx"

# Read data
AT <- read_excel(file_path)

# Filter data
AT_filtered <- AT %>%
  filter(YEAR %in% c(1990, 2014, 2018, 2020, 2022),
             LC7_text != "Natural"
  )

# Calculate proportional extent by year
AT_filtered <- AT_filtered %>%
  group_by(YEAR) %>%
  mutate(proportion = km2_total / sum(km2_total, na.rm = TRUE)) %>%
  ungroup()

# Define custom colours
custom_colors <- c(
  "Built-up" = "black",
  "Mine" = "red",
  "Artificial-waterbody" = "blue",
  "Secondary-natural" = "orange",
  "Plantation" = "purple",
  "Cropland" = "brown"
)

# Static plot (ggplot2) with line plot and no gridlines
ggplot(AT_filtered, aes(x = YEAR, y = proportion, color = LC7_text, group = LC7_text)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_color_manual(values = custom_colors) +
  labs(
    x = "Year",
    y = "Proportional extent",
    color = "Land cover class"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```

Figure 1. Trends in the proportional extent of pressures within the land cover in the Albany Thicket between 1990 and 2022.

```{r}
library(readxl)
library(dplyr)
library(ggplot2)

# -------------------------------
# 1. Read data
# -------------------------------
AT2 <- read_excel("C:/Users/K.Mogajane/OneDrive - sanbi.org.za/Kagiso Mogajane/Research Assistant 2025/LandCoverChangeAnalysis_NBA2025/R_Project/nba_landcover/data/Biome_Summaries/Albany_Thicket.xlsx")

# -------------------------------
# 2. Filter for 2022 and prepare groups
# -------------------------------
AT2_2022 <- AT2 %>%
  filter(YEAR == 2022) %>%
  mutate(
    group = ifelse(LC7_text == "Natural", "Natural", "Non-natural"),
    subgroup = ifelse(LC7_text == "Natural", "Natural", LC7_text)
  ) %>%
  group_by(group, subgroup) %>%
  summarise(km2 = sum(km2_total, na.rm = TRUE), .groups = "drop")

# -------------------------------
# 3. Custom colour scheme (hex values)
# -------------------------------
custom_colors <- c(
   "Natural" = "#2E6F40",              # green
  "Built-up" = "#000000",             # black
  "Mine" = "#FF0000",                  # red
  "Artificial-waterbody" = "#0000FF",  # blue
  "Secondary-natural" = "#FFA500",     # orange
  "Plantation" = "#800080",           # purple
  "Cropland" = "#A52A2A"                
)

# -------------------------------
# 4. Create stacked bar plot
# -------------------------------
ggplot(AT2_2022, aes(x = group, y = km2, fill = subgroup)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = custom_colors) +
  labs(
    x = "",
    y = "Area (km²)",
    fill = "Land cover type"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    text = element_text(size = 12)
  )

```

Figure 2. The extent of natural and non natural cover across the biome shows that roughly 3 500 km² of its area remains natural, while about 1 000 km² has been transformed through various land uses. These land use classes are simplified into 7 classes and including the “secondary natural” class for any pixel that transitioned from a non natural class to the natural class. This transformed portion is dominated by croplands which makes up more than half of all non-natural cover, secondary-natural, which contributes just over a quarter. Built-up areas form a noticeable share, while artificial waterbodies, plantations and mining each account for only a small fraction of the transformed landscape.

### Desert

The Desert biome occupies a small portion of the extreme northwest of the country, forming the southernmost extent of the Namib Desert. The Richtersveld region, which includes both Succulent Karoo and Desert ecosystem types, and forms part of the Succulent Karoo Global Biodiversity Hotspot @jurgens2006.

The biome experienced habitat loss between 1990 and 2022, with about 0.8% of its natural area lost (Table 1). Mining is the key pressure in this biome; however, the land cover suggests there has been a decline in extent since the 1990 dataset, likely due to abandoned mines. The expansion of built-up areas likely contributed to the loss of natural habitat as mining declined in extent.

From 1990 to 2014, mining and cropland areas declined, while secondary-natural vegetation and artificial waterbodies showed slight increases within the Desert biome (Figure 3). Plantations and built-up areas remained largely unchanged. Between 2014 and 2022, built up areas increased, while artificial waterbodies declined and secondary-natural vegetation expanded slightly. Mining continued to decline in extent. Cropland and plantation cover remained relatively unchanged.

The patterns in the cropland extent decline (Figure 3), could largely be due to water scarcity (as there is a decline in artificial waterbody extent (Figure 3), soil degradation, and limited agricultural viability in Desert ecosystems, this contraction was accompanied by a gradual expansion of secondary-natural vegetation (Figure 3), likely driven by the abandonment of farmlands and mine-closures, allowing for passive ecological recovery @theobald2015 @moyo2022. The decline of mining extent may have been driven by global fluctuations in mineral prices and regulatory tightening ultimately leading to the closure or downscaling of several mines, particularly those with shallow or exhausted deposits @lotchner2013; @theobald2025. South Africa also downsized mines, often leading to abandonment of the mining sites @nel2002; @lotchner2013; @marais2013. Through land cover data, we observe a trend where a decline in mine extent, is correlated to an increase in the increase of secondary-natural extent (Figure 3) in the same time periods with the mine closures, suggesting the abandoned mining sites may be passively re-vegetating. Built-up areas expansion around former mining towns and renewable energy zones reflects the growth of peri-urban settlements, infrastructure developments, and new energy installations such as solar farms in remote desert areas @gaeatlholwe2023.

Importantly, plantation forestry was effectively absent throughout the period, consistent with the ecological limitations of Desert biome including low soil nutrients, wind erosion, limited rooting depth, and high evapotranspiration @jurgens2006. Overall, the desert biome appears to be undergoing an ecological transition, marked by a retreat of intensive land uses and the emergence of a mosaic landscape of low-density settlement, natural vegetation recovery, and selective agricultural intensification where hydrological conditions allow. These dynamics are consistent with desertification-response models, which predict natural regrowth in previously disturbed zones under reduced human pressure @gaeatlholwe2023.

```{r}
# Load libraries
library(readxl)
library(dplyr)
library(ggplot2)

# Define file path
file_path <- "C:/Users/K.Mogajane/OneDrive - sanbi.org.za/Kagiso Mogajane/Research Assistant 2025/LandCoverChangeAnalysis_NBA2025/R_Project/nba_landcover/data/Biome_Summaries/Desert.xlsx"

# Read data
D <- read_excel(file_path)

# Filter data
D_filtered <- D %>%
  filter(YEAR %in% c(1990, 2014, 2018, 2020, 2022),
             LC7_text != "Natural"
  )

# Calculate proportional extent by year
D_filtered <- D_filtered %>%
  group_by(YEAR) %>%
  mutate(proportion = km2_total / sum(km2_total, na.rm = TRUE)) %>%
  ungroup()

# Define custom colours
custom_colors <- c(
  "Built-up" = "black",
  "Mine" = "red",
  "Artificial-waterbody" = "blue",
  "Secondary-natural" = "orange",
  "Plantation" = "purple",
  "Cropland" = "brown"
)

# Static plot (ggplot2) with line plot and no gridlines
ggplot(D_filtered, aes(x = YEAR, y = proportion, color = LC7_text, group = LC7_text)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_color_manual(values = custom_colors) +
  labs(
    x = "Year",
    y = "Proportional extent",
    color = "Land cover class"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```

Figure 3. Trends in the proportional extent of pressures within the land cover in the Desert between 1990 and 2022.

```{r}
library(readxl)
library(dplyr)
library(ggplot2)

# -------------------------------
# 1. Read data
# -------------------------------
D2 <- read_excel("C:/Users/K.Mogajane/OneDrive - sanbi.org.za/Kagiso Mogajane/Research Assistant 2025/LandCoverChangeAnalysis_NBA2025/R_Project/nba_landcover/data/Biome_Summaries/Desert.xlsx")

# -------------------------------
# 2. Filter for 2022 and prepare groups
# -------------------------------
D2_2022 <- D2 %>%
  filter(YEAR == 2022) %>%
  mutate(
    group = ifelse(LC7_text == "Natural", "Natural", "Non-natural"),
    subgroup = ifelse(LC7_text == "Natural", "Natural", LC7_text)
  ) %>%
  group_by(group, subgroup) %>%
  summarise(km2 = sum(km2_total, na.rm = TRUE), .groups = "drop")

# -------------------------------
# 3. Custom colour scheme (hex values)
# -------------------------------
custom_colors <- c(
  "Natural" = "#2E6F40",              # green
  "Built-up" = "#000000",             # black
  "Mine" = "#FF0000",                  # red
  "Artificial-waterbody" = "#0000FF",  # blue
  "Secondary-natural" = "#FFA500",     # orange
  "Plantation" = "#800080",           # purple
  "Cropland" = "#A52A2A"              # brown
)

# -------------------------------
# 4. Create stacked bar plot
# -------------------------------
ggplot(D2_2022, aes(x = group, y = km2, fill = subgroup)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = custom_colors) +
  labs(
    x = "",
    y = "Area (km²)",
    fill = "Land cover type"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    text = element_text(size = 12)
  )
```

Figure 4. The extent of natural and non natural cover across the biome shows that roughly 6 000 km² of its area remains natural, while less than 2 000 km² has been transformed through various land uses. These land use classes are simplified into 7 classes and including the “secondary natural” class for any pixel that transitioned from a non natural class to the natural class. This transformed portion is dominated by mining, the secondary-natural, built-up areas and croplands. Artificial waterbodies and plantations each account for only a small fraction of the transformed landscape.

### Forest

Indigenous forests in South Africa are small patches scattered mainly along the eastern and southern parts of the country from the coast extending inland to the escarpment of the country. However, pockets also occur in the western and northern provinces. They typically occur as small patches embedded within larger Fynbos, Grassland, Albany Thicket or Savanna biomes, though larger complexes are found in the Knysna area and in the Amathole mountain range @mucina2006c. The coastal forests of eastern South Africa make up part of the Maputaland-Pondoland-Albany Global Biodiversity Hotspot .

The Forest biome has 82% of its natural habitat remaining, with a recent natural loss rate of 0.084% per year between 1990 and 2022 (Table 2). Plantation and croplands are the key drivers of this loss (Figure 5). Mining could be an emerging pressure in this biome; land cover data indicate an increase in 2018, with mining covering a larger area within the biome.

From 1990 to 2014, land cover data indicate an inverse relationship between plantations and secondary-natural vegetation within the Forest biome, with plantation extent declining steadily while secondary-natural vegetation expanded. During this period, cropland, built-up areas, and, to a lesser extent, mining also increased, while artificial waterbodies remained stable. Between 2014 and 2022, plantations and cropland declined overall, while secondary-natural vegetation and mining increased. Built-up areas showed some fluctuation but remained generally stable, and artificial waterbodies showed little change.

Beyond these drivers, patterns of habitat loss in forest ecosystems are also linked to multiple disturbances, both natural and anthropogenic.Forest degradation and eventual habitat loss can follow stages of crown condition deterioration, where disturbance reduces canopy cover and alters species composition @geldenhuys2010. This degradation opens the forest to invasive species and pioneer plants, which can further shift the ecosystem away from mature forest conditions. A notable dynamic is the distinction between natural forest regrowth and the "naturalisation" of indigenous trees in non-forest ecosystems. Some studies explain that indigenous trees can colonise areas such as old agricultural lands, fire-suppressed grasslands, or gaps within plantations, leading to stands that may structurally resemble natural forest but differ in ecological function and origin. These "naturalised natural forests" may then be captured as secondary-natural vegetation by the land cover data as seen in Figure 5 @karumbidza2006; @geldenhuys2011; @leavercherry2020.

South Africa’s timber plantations are located either within or next to indigenous forest @karumbidza2006. Despite the consistent decline in plantations from land cover data (Figure 5), studies @karumbidza2006 ; @leavercherry2020 show that plantation extent increased in area between 1990 and 2002 by 11% in Mpumalanga, KwaZulu-Natal, and the Eastern Cape. However, in the Western Cape province, plantation forestry declined due to the decommissioning of approximately 450 km² of plantations under the Government forestry exit policy in 2001 @tiroza2016. Other pressures within the Forest biome that are not captured by the land cover, but may be classified as secondary-natural (Figure 5) due to the re-establishment of Forests @geldenhuys2011, include the harvesting of Non-Timber Forest Products (NTFP) such as bark for medicine, rope fibre, fuel wood and thatching @chirwa2008, @leavercherry2020. Harvesting for these NTFP can be highly destructive and result in increased tree mortality, which ultimately results in deforestation due to increased poverty and demand, especially in rural livelihoods @chidumayo1996; @chirwa2008; @leavercherry2020.

The natural extent of the Forest still occupies the majority of the biome at nearly 4 000 km² (82% of the biome) (Table 2 and Figure 6). However, secondary-natural vegetation and plantations remain key pressures within the Forest biome (Figure 6), driven largely by the timber and pulp industries, over-harvesting, invasive species and other disturbances leading to loss of natural habitat @karumbidza2006; @geldenhuys2010; @geldenhuys2011.

```{r}
# Load libraries
library(readxl)
library(dplyr)
library(ggplot2)

# Define file path
file_path <- "C:/Users/K.Mogajane/OneDrive - sanbi.org.za/Kagiso Mogajane/Research Assistant 2025/LandCoverChangeAnalysis_NBA2025/R_Project/nba_landcover/data/Biome_Summaries/Forest.xlsx"

# Read data
FRST <- read_excel(file_path)

# Filter data
FRST_filtered <- FRST %>%
  filter(YEAR %in% c(1990, 2014, 2018, 2020, 2022),
             LC7_text != "Natural"
  )

# Calculate proportional extent by year
FRST_filtered <- FRST_filtered %>%
  group_by(YEAR) %>%
  mutate(proportion = km2_total / sum(km2_total, na.rm = TRUE)) %>%
  ungroup()

# Define custom colours
custom_colors <- c(
  "Built-up" = "black",
  "Mine" = "red",
  "Artificial-waterbody" = "blue",
  "Secondary-natural" = "orange",
  "Plantation" = "purple",
  "Cropland" = "brown"
)

# Static plot (ggplot2) with line plot and no gridlines
ggplot(FRST_filtered, aes(x = YEAR, y = proportion, color = LC7_text, group = LC7_text)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_color_manual(values = custom_colors) +
  labs(
    x = "Year",
    y = "Proportional extent",
    color = "Land cover class"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```

Figure 5. Trends in the proportional extent of pressures within the land cover in the Forest biome between 1990 and 2022.

```{r}
library(readxl)
library(dplyr)
library(ggplot2)

# -------------------------------
# 1. Read data
# -------------------------------
FRST2 <- read_excel("C:/Users/K.Mogajane/OneDrive - sanbi.org.za/Kagiso Mogajane/Research Assistant 2025/LandCoverChangeAnalysis_NBA2025/R_Project/nba_landcover/data/Biome_Summaries/Forest.xlsx")

# -------------------------------
# 2. Filter for 2022 and prepare groups

FRST2_2022 <- FRST2 %>%
  filter(YEAR == 2022) %>%
  mutate(
    group = ifelse(LC7_text == "Natural", "Natural", "Non-natural"),
    subgroup = ifelse(LC7_text == "Natural", "Natural", LC7_text)
  ) %>%
  group_by(group, subgroup) %>%
  summarise(km2 = sum(km2_total, na.rm = TRUE), .groups = "drop")

# -------------------------------
# 3. Custom colour scheme (hex values)
# -------------------------------
custom_colors <- c(
  "Natural" = "#2E6F40",              # green
  "Built-up" = "#000000",             # black
  "Mine" = "#FF0000",                  # red
  "Artificial-waterbody" = "#0000FF",  # blue
  "Secondary-natural" = "#FFA500",     # orange
  "Plantation" = "#800080",           # purple
  "Cropland" = "#A52A2A"              # brown
)

# -------------------------------
# 4. Create stacked bar plot
# -------------------------------
ggplot(FRST2_2022, aes(x = group, y = km2, fill = subgroup)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = custom_colors) +
  labs(
    x = "",
    y = "Area (km²)",
    fill = "Land cover type"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    text = element_text(size = 12)
  )

```

Figure 6. The extent of natural and non natural cover across the biome shows that roughly 3 500 km² of its area remains natural, while about 1 000 km² has been transformed through various land uses. These land use classes are simplified into 7 classes and including the “secondary natural” class for any pixel that transitioned from a non natural class to the natural class. This transformed portion is dominated by secondary natural, which makes up more than half of all non-natural cover, croplands, which contributes just over a quarter. Built-up areas form a noticeable share, while artificial waterbodies, plantations and mining each account for only a small fraction of the transformed landscape.

### Fynbos

In this section, we will firstly explore the pressures within the Fynbos at a biome level. Then address these pressures through the Renosterveld and the Fynbos vegetation groups.

The Fynbos biome is globally recognised as a unique floral kingdom and Global Biodiversity Hotspot, with very high plant diversity and levels of endemism @rebelo2006. However, only 67% of its natural habitat remains and has lost 5.9% between 1990 and 2022 (Table 1). Multiple drivers contribute to this ongoing decline, as indicated by land cover data. Cropland emerges as the predominant pressure, having occupied the largest area within the biome since 1990. Over time cropland seems to have decreased as secondary natural area has increased, along with a proportionally smaller increase in built-up area.

A recent study @du_preez2025 assessed Indigofera genus to understand how plant species have adapted and diversified across the biome. Their findings reveal that the Fynbos biome is made up of two distinguishable vegetation groups, each shaped by unique soils, climate and fire regimes. The most important are Fynbos (found mostly on nutrient poor, sandy soil in mountain areas) and Renosterveld (found in nutrient richer, shale-based soils in lowlands). Although they often sit side by side on the landscape, these two ecosystems experience different pressures and support different plant communities. Indigofera species were as indicators of how plants have moved across these environments over millions of years @du_preez2025. Some Indigofera groups have specialised in Fynbos, especially in montane sandstone habitats where frequent fires, low nutrients and isolation drive rapid plant diversification. Other groups are adapted to the Renosterveld which is less fire-prone but highly fertile and increasingly under threat from farming and development. This evolutionary evidence from @du_preez2025 coupled with pressures unique to the Fynbos and Renosterveld vegetation groups provides justification to split our assessment of pressures in the Fynbos biome. Recognising that there are Strandveld and Shale Band vegetation groups within the Fynbos biome, studies @rebelo2006; @cowling2023 indicate a close affinity of the Strandveld and the Shale Band vegetation to Fynbos, hence we have included the ecosystem types within those vegetation groups into the Fynbos vegetation our Fynbos/Renosterveld group split.

```{r}
# Load libraries
library(readxl)
library(dplyr)
library(ggplot2)

# Define file path
file_path <- "C:/Users/K.Mogajane/OneDrive - sanbi.org.za/Kagiso Mogajane/Research Assistant 2025/LandCoverChangeAnalysis_NBA2025/R_Project/nba_landcover/data/Biome_Summaries/Fynbos.xlsx"

# Read data
FYNBS <- read_excel(file_path)

# Filter data
FYNBS_filtered <- FYNBS %>%
  filter(YEAR %in% c(1990, 2014, 2018, 2020, 2022),
             LC7_text != "Natural"
  )

# Calculate proportional extent by year
FYNBS_filtered <- FYNBS_filtered %>%
  group_by(YEAR) %>%
  mutate(proportion = km2_total / sum(km2_total, na.rm = TRUE)) %>%
  ungroup()

# Define custom colours
custom_colors <- c(
  "Built-up" = "black",
  "Mine" = "red",
  "Artificial-waterbody" = "blue",
  "Secondary-natural" = "orange",
  "Plantation" = "purple",
  "Cropland" = "brown"
)

# Static plot (ggplot2) with line plot and no gridlines
ggplot(FYNBS_filtered, aes(x = YEAR, y = proportion, color = LC7_text, group = LC7_text)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_color_manual(values = custom_colors) +
  labs(
    x = "Year",
    y = "Proportional extent",
    color = "Land cover class"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```

Figure 7. The trends in land cover change for the Fynbos biome between 1990 and 2022.

```{r}
library(readxl)
library(dplyr)
library(ggplot2)

# -------------------------------
# 1. Read data
# -------------------------------
FRST2 <- read_excel("C:/Users/K.Mogajane/OneDrive - sanbi.org.za/Kagiso Mogajane/Research Assistant 2025/LandCoverChangeAnalysis_NBA2025/R_Project/nba_landcover/data/Biome_Summaries/Fynbos.xlsx")

# -------------------------------
# 2. Filter for 2022 and prepare groups

FRST2_2022 <- FRST2 %>%
  filter(YEAR == 2022) %>%
  mutate(
    group = ifelse(LC7_text == "Natural", "Natural", "Non-natural"),
    subgroup = ifelse(LC7_text == "Natural", "Natural", LC7_text)
  ) %>%
  group_by(group, subgroup) %>%
  summarise(km2 = sum(km2_total, na.rm = TRUE), .groups = "drop")

# -------------------------------
# 3. Custom colour scheme (hex values)
# -------------------------------
custom_colors <- c(
  "Natural" = "#2E6F40",              # green
  "Built-up" = "#000000",             # black
  "Mine" = "#FF0000",                  # red
  "Artificial-waterbody" = "#0000FF",  # blue
  "Secondary-natural" = "#FFA500",     # orange
  "Plantation" = "#800080",           # purple
  "Cropland" = "#A52A2A"              # brown
)

# -------------------------------
# 4. Create stacked bar plot
# -------------------------------
ggplot(FRST2_2022, aes(x = group, y = km2, fill = subgroup)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = custom_colors) +
  labs(
    x = "",
    y = "Area (km²)",
    fill = "Land cover type"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    text = element_text(size = 12)
  )

```

Figure 7b. The extent of natural and non natural cover across the biome shows that roughly 50 00 km² of its area remains natural, while about 30 000 km² has been transformed through various land uses. These land use classes are simplified into 7 classes and including the “secondary natural” class for any pixel that transitioned from a non natural class to the natural class. This transformed portion is dominated by cropland, which makes up more than half of all non-natural cover, then secondary-natural, which contributes just over a quarter. Built-up and plantations areas form a noticeable share, while artificial waterbodies and mining each account for only a small fraction of the transformed landscape.

**Fynbos vegetation**

Here we include the Strandveld and Shale Band Vegetation within the larger Fynbos subgroup due to the close plant communities and geological and soil associations @rebelo2006; @cowling2023.

Fynbos vegetation is dominated by hard and fine-leaved shrubs that are adapted to nutrient-poor, acidic sandy soils @rebelo2006. Within this subgroup, Shale Band vegetation is found in lower slopes and valleys, while Strandveld occurs on alkaline, calcareous coastal sands and supports taller evergreen shrubs and succulents @rebelo2006.

Within the Fynbos vegetation, croplands, by far the most extensive land cover, declined in extent from 1990 to 2014, along with plantations and artificial waterbodies. During the same period, built-up areas and secondary-natural vegetation increased, while mining showed little to no change. Between 2014 and 2022, cropland and artificial waterbody extent fluctuated but generally declined. Built-up areas and secondary-natural vegetation continued to expand, and plantations showed an overall increase. Mining remained largely unchanged throughout.

Habitat transformation and fragmentation are a major threat to lowland ecosystems in the Fynbos Biome @ntshanga2021, driven primarily by urban expansion into lowland Fynbos and Strandveld around Cape Town and coastal areas. This transformation has led to the isolation of remnant natural patches within heavily altered, human-dominated landscapes. In parallel, post-disturbance vegetation dynamics in the biome have contributed to shifts in land cover classifications. The removal of invasive alien plants, particularly through spontaneous succession, can result in vegetation states dominated by native species, yet functionally and compositionally distinct from original fynbos @holmes2020. These shifts are often captured in land cover data sets as secondary natural vegetation, particularly where plantation cover declines following alien clearance. Ecological thresholds are crossed when prolonged invasion alters soil or suppresses key fynbos shrubs, leading to transitions toward novel fynbos community assemblages. Long-term restoration in fynbos shrublands does not always recover original biodiversity or structure, despite alien plant removal @fill2017. Collectively, these studies @fill2017 @holmes2020 illustrate how recovery trajectories, even when native-dominated, often produce altered states that contribute to increases in the secondary natural class while reflecting a legacy of ecological disturbance.

In Figure 10, the built-up, Studies showed that Fynbos vegetation faces threats of habitat loss and fragmentation from urban development, agriculture, plantation forestry and the spread of invasive alien plants @sandberg2016.

```{r}
# Load libraries
library(readxl)
library(dplyr)
library(ggplot2)

# Define file path
file_path <- "C:/Users/K.Mogajane/OneDrive - sanbi.org.za/Kagiso Mogajane/Research Assistant 2025/LandCoverChangeAnalysis_NBA2025/R_Project/nba_landcover/data/Biome_Summaries/FSSBand_VG.xlsx"

# Read data
FSSBand_VG <- read_excel(file_path)

# Filter data
FSSBand_VG_filtered <- FSSBand_VG %>%
  filter(YEAR %in% c(1990, 2014, 2018, 2020, 2022),
             LC7_text != "Natural"
  )

# Calculate proportional extent by year
FSSBand_VG_filtered <- FSSBand_VG_filtered %>%
  group_by(YEAR) %>%
  mutate(proportion = km2_total / sum(km2_total, na.rm = TRUE)) %>%
  ungroup()

# Define custom colours
custom_colors <- c(
  "Built-up" = "black",
  "Mine" = "red",
  "Artificial-waterbody" = "blue",
  "Secondary-natural" = "orange",
  "Plantation" = "purple",
  "Cropland" = "brown"
)

# Static plot (ggplot2) with line plot and no gridlines
ggplot(FSSBand_VG_filtered, aes(x = YEAR, y = proportion, color = LC7_text, group = LC7_text)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_color_manual(values = custom_colors) +
  labs(
    x = "Year",
    y = "Proportional extent",
    color = "Land cover class"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
 
```

Figure 8. Trends in land cover change within the Fynbos vegetation group between 1990 and 2022.

```{r}
library(readxl)
library(dplyr)
library(ggplot2)

# -------------------------------
# 1. Read data
# -------------------------------
FSSBand_VG2 <- read_excel("C:/Users/K.Mogajane/OneDrive - sanbi.org.za/Kagiso Mogajane/Research Assistant 2025/LandCoverChangeAnalysis_NBA2025/R_Project/nba_landcover/data/Biome_Summaries/FSSBand_VG.xlsx")

# -------------------------------
# 2. Filter for 2022 and prepare groups

FSSBand_VG2_2022 <- FSSBand_VG2 %>%
  filter(YEAR == 2022) %>%
  mutate(
    group = ifelse(LC7_text == "Natural", "Natural", "Non-natural"),
    subgroup = ifelse(LC7_text == "Natural", "Natural", LC7_text)
  ) %>%
  group_by(group, subgroup) %>%
  summarise(km2 = sum(km2_total, na.rm = TRUE), .groups = "drop")

# -------------------------------
# 3. Custom colour scheme (hex values)
# -------------------------------
custom_colors <- c(
  "Natural" = "#2E6F40",              # green
  "Built-up" = "#000000",             # black
  "Mine" = "#FF0000",                  # red
  "Artificial-waterbody" = "#0000FF",  # blue
  "Secondary-natural" = "#FFA500",     # orange
  "Plantation" = "#800080",           # purple
  "Cropland" = "#A52A2A"              # brown
)

# -------------------------------
# 4. Create stacked bar plot
# -------------------------------
ggplot(FSSBand_VG2_2022, aes(x = group, y = km2, fill = subgroup)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = custom_colors) +
  labs(
    x = "",
    y = "Area (km²)",
    fill = "Land cover type"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    text = element_text(size = 12)
  )
```

Figure 9. These land use classes are simplified into 7 classes and including the “secondary natural” class for any pixel that transitioned from a non natural class to the natural class. This transformed portion is dominated by cropland, which makes up more than half of all non-natural cover, then secondary-natural, which contributes just over a quarter. Built-up areas and plantations form a noticeable share, while artificial waterbodies and mining each account for only a small fraction of the transformed landscape.

**Renosterveld vegetation**

Renosterveld vegetation is is characterized by a predominantly evergreen shrubs with a mix of grasses. It primarily occurs in the fertile lowlands within the Cape Floristic Region @rebelo2006; @cousins2018.

Between 1990 and 2014, cropland, plantation, and artificial waterbodies declined in extent within the Renosterveld, while secondary-natural areas increased notably largely due to decommissioning of the of approximately 45 000 hectares of plantation forestry by the Government forestry exit policy in 2001 @tiroza2016. Built-up and mining cover remained largely unchanged. From 2014 to 2022, cropland initially declined before expanding again, while plantation and artificial waterbodies showed minor increases. Secondary-natural and built-up areas fluctuated over time. Mining extent remained relatively stable throughout.

@ntshanga2021 examined patterns of habitat fragmentation in the Fynbos biome and noted significant transformation of the Renosterveld vegetation for agriculture, particularly in lowland areas. This has resulted in a landscape dominated by fragmented remnants embedded within cultivated lands. @kemper1999 explored the ecological consequences of this fragmentation by assessing plant community structure in Renosterveld fragments. While overall species richness did not show a consistent relationship with fragment size, larger fragments were more similar in community composition than smaller ones, suggesting a decline in ecological integrity with increasing fragmentation. These findings reflect a broader pattern in which remaining Renosterveld fragments often persist in a secondary natural state. The variation observed across fragment sizes supports the notion that many of these areas represent ecologically altered states, shaped by historical land-use change and ongoing edge effects, which are captured in land cover data as secondary natural vegetation.

Drought within the Western Cape between 2015 and 2018 @pienaar2018; @mahlalela2020; @orimoloye2022 resulted in a decline in crop production, with a 14% decline across the agricultural industry within the Western Cape province particularly in the Renosterveld. This was due to reduced access to water resources @pienaar2018. These effects are couple with the decline in artificial waterbody and cropland decline in the 2014 and 2018 in the land cover data (Figure 10).

```{r}
# Load libraries
library(readxl)
library(dplyr)
library(ggplot2)

# Define file path
file_path <- "C:/Users/K.Mogajane/OneDrive - sanbi.org.za/Kagiso Mogajane/Research Assistant 2025/LandCoverChangeAnalysis_NBA2025/R_Project/nba_landcover/data/Biome_Summaries/Renosterveld_VG.xlsx"

# Read data
Renosterveld_VG <- read_excel(file_path)

# Filter data
Renosterveld_VG_filtered <- Renosterveld_VG %>%
  filter(YEAR %in% c(1990, 2014, 2018, 2020, 2022),
             LC7_text != "Natural"
  )

# Calculate proportional extent by year
Renosterveld_VG_filtered <- Renosterveld_VG_filtered %>%
  group_by(YEAR) %>%
  mutate(proportion = km2_total / sum(km2_total, na.rm = TRUE)) %>%
  ungroup()

# Define custom colours
custom_colors <- c(
  "Built-up" = "black",
  "Mine" = "red",
  "Artificial-waterbody" = "blue",
  "Secondary-natural" = "orange",
  "Plantation" = "purple",
  "Cropland" = "brown"
)

# Static plot (ggplot2) with line plot and no gridlines
ggplot(Renosterveld_VG_filtered, aes(x = YEAR, y = proportion, color = LC7_text, group = LC7_text)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_color_manual(values = custom_colors) +
  labs(
    x = "Year",
    y = "Proportional extent",
    color = "Land cover class"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```

Figure 10. The proportional extent of pressures within the Renosterveld vegetation between 1990 and 2022 land cover sets.

```{r}
library(readxl)
library(dplyr)
library(ggplot2)

# -------------------------------
# 1. Read data
# -------------------------------
Renosterveld_VG2 <- read_excel("C:/Users/K.Mogajane/OneDrive - sanbi.org.za/Kagiso Mogajane/Research Assistant 2025/LandCoverChangeAnalysis_NBA2025/R_Project/nba_landcover/data/Biome_Summaries/Renosterveld_VG.xlsx")

# -------------------------------
# 2. Filter for 2022 and prepare groups

Renosterveld_VG2_2022 <- Renosterveld_VG2 %>%
  filter(YEAR == 2022) %>%
  mutate(
    group = ifelse(LC7_text == "Natural", "Natural", "Non-natural"),
    subgroup = ifelse(LC7_text == "Natural", "Natural", LC7_text)
  ) %>%
  group_by(group, subgroup) %>%
  summarise(km2 = sum(km2_total, na.rm = TRUE), .groups = "drop")

# -------------------------------
# 3. Custom colour scheme (hex values)
# -------------------------------
custom_colors <- c(
  "Natural" = "#2E6F40",              # green
  "Built-up" = "#000000",             # black
  "Mine" = "#FF0000",                  # red
  "Artificial-waterbody" = "#0000FF",  # blue
  "Secondary-natural" = "#FFA500",     # orange
  "Plantation" = "#800080",           # purple
  "Cropland" = "#A52A2A"              # brown
)

# -------------------------------
# 4. Create stacked bar plot
# -------------------------------
ggplot(Renosterveld_VG2_2022, aes(x = group, y = km2, fill = subgroup)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = custom_colors) +
  labs(
    x = "",
    y = "Area (km²)",
    fill = "Land cover type"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    text = element_text(size = 12)
  )
```

Figure 11. These land use classes are simplified into 7 classes and including the “secondary natural” class for any pixel that transitioned from a non natural class to the natural class. This transformed portion is dominated by cropland, which makes up more than half of all non-natural cover, then secondary-natural, which contributes just over a quarter. Built-up areas form a noticeable share, while artificial waterbodies, plantations and mining each account for only a small fraction of the transformed landscape.

### Indian Ocean Coastal Belt

The Indian Ocean Coastal Belt biome forms part of the Maputaland-Pondoland-Albany Global Biodiversity Hotspot and is a heterogenous complex of forest, mesic savanna and grassland vegetation communities; unique both ecologically and biogeographically. This biome’s range extends across the entire KwaZulu-Natal coast and small parts of the Eastern Cape’s northeastern coast @mucina2006b.

The Indian Ocean Coastal Belt has the least remaining natural habitat of all the biomes at 33% natural habitat remaining as of the 2022 land cover data set. It has the highest rate of natural habitat lost for all biomes, losing 0.66% natural cover per year (Table 1). The land cover suggests that this loss of natural habitat is due to the high extent of built-up areas, croplands and plantations (Figure 11).

From 1990 to 2014, built-up areas and plantations declined noticeably, while croplands and secondary-natural vegetation increased considerably. Mining and artificial waterbodies remained mostly stable, with a slight increase in artificial waterbodies. Between 2014 and 2022, cropland and secondary-natural vegetation initially increased but later declined, while plantation cover increased overall. Built-up areas fluctuated, showing a general decline by 2022. Mining and artificial waterbodies remained largely unchanged throughout.

Coastal areas of the KwaZulu-Natal province are densely populated, and as a result, there are numerous economic activities that that result in the alteration of natural habitat @goble2022; @kznedtea2024. There have been increasing developments of residential areas and port facilities along the coast @goble2022, these developments contribute to habitat loss and the land cover has captured this, hence the increase in the extent of built-up areas (Figure 12).

South Africa’s timber plantations are located either within or next to indigenous forest @karumbidza2006 and sometimes within the Indian Ocean Coastal Belt. For example, in KwaZulu-Natal, in the Richard’s Bay area, there are industrial timber plantation that are where Indian Ocean Coastal Belt has been mapped in the National Vegetation Map @mucina2006. Conversion of natural land to monoculture plantation is another driver of habitat loss @karumbidza2006. Industrial timber plantations seem to be infested with alien invasive trees such as Black wattle (*Acacia mearnsii*) @karumbidza2006. This invasion is because in commercial plantations, some alien trees species are used but they end up invading event areas where they were not meant to occupy @dewit2001 @karumbidza2006. The national invasive trees survey shows that the Indian Ocean Coastal Belt is the most alien invaded biome in the country @kotze2025.

```{r}
# Load libraries
library(readxl)
library(dplyr)
library(ggplot2)

# Define file path
file_path <- "C:/Users/K.Mogajane/OneDrive - sanbi.org.za/Kagiso Mogajane/Research Assistant 2025/LandCoverChangeAnalysis_NBA2025/R_Project/nba_landcover/data/Biome_Summaries/IOCB.xlsx"

# Read data
IOCB <- read_excel(file_path)

# Filter data
IOCB_filtered <- IOCB %>%
  filter(YEAR %in% c(1990, 2014, 2018, 2020, 2022),
             LC7_text != "Natural"
  )

# Calculate proportional extent by year
IOCB_filtered <- IOCB_filtered %>%
  group_by(YEAR) %>%
  mutate(proportion = km2_total / sum(km2_total, na.rm = TRUE)) %>%
  ungroup()

# Define custom colours
custom_colors <- c(
  "Built-up" = "black",
  "Mine" = "red",
  "Artificial-waterbody" = "blue",
  "Secondary-natural" = "orange",
  "Plantation" = "purple",
  "Cropland" = "brown"
)

# Static plot (ggplot2) with line plot and no gridlines
ggplot(IOCB_filtered, aes(x = YEAR, y = proportion, color = LC7_text, group = LC7_text)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_color_manual(values = custom_colors) +
  labs(
    x = "Year",
    y = "Proportional extent",
    color = "Land cover class"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```

Figure 12. The proportional extent of pressures within the Indian Ocean Coastal Belt biome between 1990 and 2022 land cover sets.

```{r}
library(readxl)
library(dplyr)
library(ggplot2)

# -------------------------------
# 1. Read data
# -------------------------------
IOCB2 <- read_excel("C:/Users/K.Mogajane/OneDrive - sanbi.org.za/Kagiso Mogajane/Research Assistant 2025/LandCoverChangeAnalysis_NBA2025/R_Project/nba_landcover/data/Biome_Summaries/IOCB.xlsx")

# -------------------------------
# 2. Filter for 2022 and prepare groups

IOCB2_2022 <- IOCB2 %>%
  filter(YEAR == 2022) %>%
  mutate(
    group = ifelse(LC7_text == "Natural", "Natural", "Non-natural"),
    subgroup = ifelse(LC7_text == "Natural", "Natural", LC7_text)
  ) %>%
  group_by(group, subgroup) %>%
  summarise(km2 = sum(km2_total, na.rm = TRUE), .groups = "drop")

# -------------------------------
# 3. Custom colour scheme (hex values)
# -------------------------------
custom_colors <- c(
  "Natural" = "#2E6F40",              # green
  "Built-up" = "#000000",             # black
  "Mine" = "#FF0000",                  # red
  "Artificial-waterbody" = "#0000FF",  # blue
  "Secondary-natural" = "#FFA500",     # orange
  "Plantation" = "#800080",           # purple
  "Cropland" = "#A52A2A"              # brown
)

# -------------------------------
# 4. Create stacked bar plot
# -------------------------------
ggplot(IOCB2_2022, aes(x = group, y = km2, fill = subgroup)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = custom_colors) +
  labs(
    x = "",
    y = "Area (km²)",
    fill = "Land cover type"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    text = element_text(size = 12)
  )
```

Figure 13. The extent of natural and non natural cover across the biome shows that roughly 4 000 km² of its area remains natural, while about 8 000 km² has been transformed through various land uses. There is more transformed areas than there are natural areas within the biome. These land use classes are simplified into 7 classes and including the “secondary natural” class for any pixel that transitioned from a non natural class to the natural class. This transformed portion is dominated by built-up areas, followed by croplands, and then secondary-natural. Plantations form a noticeable share, while artificial waterbodies and mining each account for only a small fraction of the transformed landscape.

### Grassland

The Grassland biome occupies the central plateau of South Africa (Highveld) and lower elevation portions of the Eastern Cape and KZN interior. Most of the biome is dominated by C4 grasses, with C3 dominated grasslands restricted to montane areas @mucina2006d . As seen in Table 1, 8.5% of the natural habitat has been lost between 1990 and 2022. The land cover suggests that agricultural activity such as cropland and plantation are the main drivers of this loss of habitat. In addition, infrastructure is another contributor to the habitat loss as built-up areas also have a high extent within the biome and have been on a consistent rise since 1990 (Figure 14).

Between 1990 and 2014, cropland and built-up areas declined slightly, while secondary-natural vegetation increased. Plantation, artificial waterbodies, and mining remained stable. From 2014 to 2018, cropland, artificial waterbodies, and mining declined, while secondary-natural, built-up, and plantation areas expanded. Between 2018 and 2020, cropland increased slightly, while secondary-natural, artificial waterbodies, and mining declined. Built-up and plantation remained unchanged. From 2020 to 2022, cropland and secondary-natural areas declined, plantation and artificial waterbodies increased, and built-up and mining remained stable.

Habitat loss in grassland ecosystems is strongly driven by multiple land use pressures, with croplands exerting the most extensive impact due to the dominance of the agricultural sector @slooten2023; @skowno2021a, @skowno2021b. Grasslands are also extensively used for grazing by livestock and game but increasing agricultural expansion and intensification have intensified landscape fragmentation @slooten2023. In addition to these pressures, severe droughts in 2015 and 2018 likely forced a pause in cropping, reflected in declining cropland, while water shortages during these years also likely dried farm dams and reservoirs, driving a drop in artificial waterbodies @orimoloye2022. Between 2018 and 2020, cropland recovered slightly as farming resumed @orimoloye2022, but waterbodies continued to decline, indicating slow hydrological recovery.

@niemandt2016 documented that pressures from timber plantations, developments for human settlements, and clearing land for potato and kiwifruit crops in the critically endangered Woodbush Granite Grassland (WGG), where timber contributed to an annual habitat loss rate of 3.1% between 1948 and 2008, increasing patch isolation and reducing patch size. Although the WGG is endangered and faces multiple pressures, some native species populations such as Aloes lettyae are stable in the natural habitats of the remnant patches of the ecosystem @kremer-kohne2020. @muller2021 further demonstrated that disturbance causes severe declines in native floristic diversity, with a 92% reduction in forb species richness and a 27% overall loss in plant diversity. Their findings also show a compositional shift toward disturbance-adapted species, though no evidence of widespread woody encroachment was found. Together, these studies highlight how land use change, especially through croplands, grazing, and plantation development, has driven direct transformation and fragmentation, shaping the observed patterns of cropland, plantation, and secondary-natural classes in Figure 13.

Urban expansion also plays a significant role in habitat loss, particularly in regions such as Gauteng. Between 1986 and 2019, 553.23 km² (11.7%) of Critical Biodiversity Areas were converted to agricultural use and 209.70 km² (4.4%) to urban development @nhamo2021 . Ecological Support Areas experienced even greater transformation, with 972.44 km² (29.2%) lost to agriculture and 354.76 km² (10.7%) to urban land use. These figures highlight the scale of natural habitat loss resulting from expanding built-up areas, adding to the cumulative pressure already exerted by agriculture and plantations, and contributing to habitat loss. Farms’s performances in parts the North West province was assessed based on their outputs, to assess the performance and progress of land reform where farms were redistributed to historical marginalised people @kirsten2016. A comparison was done between 2005 and 2010 performance and outputs and results indicated a decline in both categories which was attributed to many factors including limited resources to invest in and maintain infrastructure which ultimately led to not farming @kirsten2016. Consequently, this may have allowed revegetation within the abandoned farms @moyo2022. Such studies @kirsten2016; @moyo2022 help us understand the factors driving the increase in secondary-natural extent.

Although low in proportional extent, it is worth noting that the highveld region (mainly Mpumalanga province) of the Grassland has numerous of mines which include 12-coal powered fire stations and at least 70 coal mines @kamolane-kgadima2024. In addition to driving habitat loss, these mines are reported to being the highest contributors to air pollution in within the region @kamolane-kgadima2024. Other highveld regions include the Gauteng, Free State and North West provinces where mining also occurs and has contributed to habitat loss, water and air pollution due to acid mine drainage, particularly in Gauteng @minnaar2020.

```{r}
# Load libraries
library(readxl)
library(dplyr)
library(ggplot2)

# Define file path
file_path <- "C:/Users/K.Mogajane/OneDrive - sanbi.org.za/Kagiso Mogajane/Research Assistant 2025/LandCoverChangeAnalysis_NBA2025/R_Project/nba_landcover/data/Biome_Summaries/Grassland.xlsx"

# Read data
GL <- read_excel(file_path)

# Filter data
GL_filtered <- GL %>%
  filter(YEAR %in% c(1990, 2014, 2018, 2020, 2022),
             LC7_text != "Natural"
  )

# Calculate proportional extent by year
GL_filtered <- GL_filtered %>%
  group_by(YEAR) %>%
  mutate(proportion = km2_total / sum(km2_total, na.rm = TRUE)) %>%
  ungroup()

# Define custom colours
custom_colors <- c(
  "Built-up" = "black",
  "Mine" = "red",
  "Artificial-waterbody" = "blue",
  "Secondary-natural" = "orange",
  "Plantation" = "purple",
  "Cropland" = "brown"
)

# Static plot (ggplot2) with line plot and no gridlines
ggplot(GL_filtered, aes(x = YEAR, y = proportion, color = LC7_text, group = LC7_text)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_color_manual(values = custom_colors) +
  labs(
    x = "Year",
    y = "Proportional extent",
    color = "Land cover class"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```

Figure 14. Trends in land cover changes between 1990 and 2022 for the Grassland biome.

```{r}
library(readxl)
library(dplyr)
library(ggplot2)

# -------------------------------
# 1. Read data
# -------------------------------
GL2 <- read_excel("C:/Users/K.Mogajane/OneDrive - sanbi.org.za/Kagiso Mogajane/Research Assistant 2025/LandCoverChangeAnalysis_NBA2025/R_Project/nba_landcover/data/Biome_Summaries/Grassland.xlsx")

# -------------------------------
# 2. Filter for 2022 and prepare groups

GL2_2022 <- GL2 %>%
  filter(YEAR == 2022) %>%
  mutate(
    group = ifelse(LC7_text == "Natural", "Natural", "Non-natural"),
    subgroup = ifelse(LC7_text == "Natural", "Natural", LC7_text)
  ) %>%
  group_by(group, subgroup) %>%
  summarise(km2 = sum(km2_total, na.rm = TRUE), .groups = "drop")

# -------------------------------
# 3. Custom colour scheme (hex values)
# -------------------------------
custom_colors <- c(
  "Natural" = "#2E6F40",              # green
  "Built-up" = "#000000",             # black
  "Mine" = "#FF0000",                  # red
  "Artificial-waterbody" = "#0000FF",  # blue
  "Secondary-natural" = "#FFA500",     # orange
  "Plantation" = "#800080",           # purple
  "Cropland" = "#A52A2A"              # brown
)

# -------------------------------
# 4. Create stacked bar plot
# -------------------------------
ggplot(GL2_2022, aes(x = group, y = km2, fill = subgroup)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = custom_colors) +
  labs(
    x = "",
    y = "Area (km²)",
    fill = "Land cover type"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    text = element_text(size = 12)
  )
```

Figure 15. The extent of natural and non natural cover across the biome shows that roughly 200 000 km² of its area remains natural, while about 140 000 km² has been transformed through various land uses. There is more transformed areas than there are natural areas within the biome. These land use classes are simplified into 7 classes and including the “secondary natural” class for any pixel that transitioned from a non natural class to the natural class. This transformed portion is dominated by cropland, followed by secondary-natural and then built-up areas. Plantations form a noticeable share, while artificial waterbodies and mining each account for only a small fraction of the transformed landscape.

### Nama-Karoo

The Nama-Karoo lies to the west of the Grassland biome and makes up much of the arid interior of the country. It receives summer rainfall and is dominated by drought tolerant shrubs and C3 grasses @mucina2006e. The Nama-Karoo is largely natural, with 0.6% of its natural habitat lost between 1990 and 2022 (Table 1). However, regardless of the small percentage loss, there are pressures impacting the biome. Land cover data suggests that since 1990, croplands have been the most impactful in this biome and continue to be, although they have reduced in extent (Figure 16). An emerging pressure that is increasing in impact is infrastructure due to the rise of built-up areas recorded by the land cover. They have especially increase since the 2014 land cover data set.

Between 1990 and 2014, the extent of cropland, artificial waterbodies, built-up areas, and mining declined, while secondary-natural vegetation increased and plantation remained stable. From 2014 to 2022, cropland and mining declined overall, with some fluctuations. Secondary-natural vegetation and artificial waterbodies showed general increases, while built-up areas declined. Plantation cover remained relatively stable throughout. Although the Nama-Karoo appears ecologically stable at a biome scale, this masks significant shifts in how the land is being used. Historically, the biome was transformed through a combination of colonial-era livestock overstocking, widespread fencing, and water abstraction, which collectively disrupted the ecological processes that had previously maintained resilience in this arid system @hoffman2018; @walker2018; @milton2021. Overgrazing from the 18th century onward led to significant shifts in vegetation structure, including the long-term loss of palatable grasses and shrubs, and a corresponding increase in unpalatable and invasive species @milton2021.

While cropland has declined in overall extent, it continues to exert the most pressure due to historical cultivation legacies. Fields abandoned after peak cultivation in the mid-20th century are now gradually reverting to secondary-natural vegetation @hoffman2018. This trend aligns with the observed increases in this cover type across several time intervals. However, a notable shift since 2014 has been the expansion of built infrastructure, particularly in the form of large-scale scientific and energy projects. The Square Kilometre Array (SKA), a major international radio astronomy installation near Carnarvon, has transformed over 130,000 hectares into a science-focused reserve with strict land use controls @hoffman2018. Simultaneously, wind and solar energy developments have proliferated across the Upper Karoo and Bushmanland bioregions, areas that serve as ecologically important flat plains for seasonal movement of wildlife and connectivity between intact habitats @hoffman2018; @gaeatlholwe2023. Threats posed by shale gas prospecting in this biome include reduced water availability @de_kock2017.

These emerging pressures overlap with short-term fluctuations in land cover categories, such as intermittent increases in artificial waterbodies and croplands. Such patterns likely reflect temporary responses to rainfall variation, economic cycles, or energy development phases @walker2018. For instance, croplands may recover during wetter periods. @walker2018 emphasize that governance decisions, development priorities, and climate shifts increasingly shape land use in the Nama-Karoo, often in ways that outpace ecological recovery or institutional planning. The transformation of plains, drainage lines, and bioregional linkages particularly in the Upper Karoo, Lower Karoo, and Bushmanland; represents more than isolated land cover change; these are the corridors that support ecological flow across a biome already limited in formal protection @hoffman2018. The integration of large-scale projects into these areas introduces new socio-ecological dynamics that are not only spatially extensive but also structurally transformative @walker2018.

```{r}
# Load libraries
library(readxl)
library(dplyr)
library(ggplot2)

# Define file path
file_path <- "C:/Users/K.Mogajane/OneDrive - sanbi.org.za/Kagiso Mogajane/Research Assistant 2025/LandCoverChangeAnalysis_NBA2025/R_Project/nba_landcover/data/Biome_Summaries/Nama-Karoo.xlsx"

# Read data
NK <- read_excel(file_path)

# Filter data
NK_filtered <- NK %>%
  filter(YEAR %in% c(1990, 2014, 2018, 2020, 2022),
             LC7_text != "Natural"
  )

# Calculate proportional extent by year
NK_filtered <- NK_filtered %>%
  group_by(YEAR) %>%
  mutate(proportion = km2_total / sum(km2_total, na.rm = TRUE)) %>%
  ungroup()

# Define custom colours
custom_colors <- c(
  "Built-up" = "black",
  "Mine" = "red",
  "Artificial-waterbody" = "blue",
  "Secondary-natural" = "orange",
  "Plantation" = "purple",
  "Cropland" = "brown"
)

# Static plot (ggplot2) with line plot and no gridlines
ggplot(NK_filtered, aes(x = YEAR, y = proportion, color = LC7_text, group = LC7_text)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_color_manual(values = custom_colors) +
  labs(
    x = "Year",
    y = "Proportional extent",
    color = "Land cover class"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```

Figure 16. The proportional extent of pressures within the Nama-Karoo biome between 1990 and 2022 land cover sets.

```{r}
library(readxl)
library(dplyr)
library(ggplot2)

# -------------------------------
# 1. Read data
# -------------------------------
NK2 <- read_excel("C:/Users/K.Mogajane/OneDrive - sanbi.org.za/Kagiso Mogajane/Research Assistant 2025/LandCoverChangeAnalysis_NBA2025/R_Project/nba_landcover/data/Biome_Summaries/Nama-Karoo.xlsx")

# -------------------------------
# 2. Filter for 2022 and prepare groups

NK2_2022 <- NK2 %>%
  filter(YEAR == 2022) %>%
  mutate(
    group = ifelse(LC7_text == "Natural", "Natural", "Non-natural"),
    subgroup = ifelse(LC7_text == "Natural", "Natural", LC7_text)
  ) %>%
  group_by(group, subgroup) %>%
  summarise(km2 = sum(km2_total, na.rm = TRUE), .groups = "drop")

# -------------------------------
# 3. Custom colour scheme (hex values)
# -------------------------------
custom_colors <- c(
  "Natural" = "#2E6F40",              # green
  "Built-up" = "#000000",             # black
  "Mine" = "#FF0000",                  # red
  "Artificial-waterbody" = "#0000FF",  # blue
  "Secondary-natural" = "#FFA500",     # orange
  "Plantation" = "#800080",           # purple
  "Cropland" = "#A52A2A"              # brown
)

# -------------------------------
# 4. Create stacked bar plot
# -------------------------------
ggplot(NK2_2022, aes(x = group, y = km2, fill = subgroup)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = custom_colors) +
  labs(
    x = "",
    y = "Area (km²)",
    fill = "Land cover type"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    text = element_text(size = 12)
  )
```

Figure 17. The extent of natural and non natural cover across the biome shows that roughly 240 000 km² of its area remains natural, while less than 25 000 km² has been transformed through various land uses. There is more transformed areas than there are natural areas within the biome. These land use classes are simplified into 7 classes and including the “secondary natural” class for any pixel that transitioned from a non natural class to the natural class. This transformed portion is dominated by cropland, followed by secondary-natural and then built-up areas. Plantations form a noticeable share, while artificial waterbodies and mining each account for only a small fraction of the transformed landscape.

### Succulent Karoo

The Succulent Karoo biome is primarily located to the west of the western escarpment, extending along the western regions of the Northern and Western Cape provinces. It is a globally Biodiversity Hotspot characterised by exceptional succulent plant diversity. This biome is largely in tact with a natural habitat loss of 1% between 1990 and 2022 (Table 1). However, there are land use pressures impacting this biome. Although croplands have declined in extent in recent years, reflected in the 2018 land cover, they are still a major driver of loss. Behind cropland is mining which has remained relatively stable since 1990 and had minor change (Figure 18).

Between 1990 and 2014, the extent of cropland, mining, and artificial waterbodies declined, while secondary-natural vegetation increased. Built-up areas and plantations remained stable. From 2014 to 2022, cropland showed an overall decline with some fluctuation, while artificial waterbodies and secondary-natural vegetation generally increased. Built-up areas and mining also increased slightly over this period, and plantation extent remained unchanged.

Many parts of the Succulent karoo are undergoing extensive grazing, coupled with croplands and mining @jonas2004 . These findings show that these land uses contribute to the loss of biodiversity and natural habitat. Different grazing intensities reduce plant diversity and endemic species in intensive grazing sites and in moderate grazing sites some endemic species thrived @haarmeyer2010. A shift in plant communities due to heavy grazing favours hardy, non-native species over unique endemics and that reduced seedling survival limits future plant regeneration within the area @haarmeyer2010. The increasing trend of the secondary-natural class then does not suggest thatall native species regenerate after the disturbances.

The Okiep District in Namaqualand was mined at multiple localities for copper for decades since the mid-1800s.Mining productivity began to decline rapidly from the mid-1990s and eventually ceased in 2007 @rozendaal2013. Since the closure of the mines, the area has been under rehabilitation @rozendaal2013. Long-term environmental monitoring studies in Namaqualand provide further context to these patterns (Figure 18). @davis2017 used 30 years of satellite data to detect a significant extension of the growing season across many areas of Namaqualand, driven by shifts in phenology (there was an earlier start and later end to the season). This expansion of the growing period, particularly in the north of Namaqualand, is linked to a combination of factors including climate warming and land-use changes. The analysis found increases in vegetation productivity, especially in previously overgrazed or abandoned areas. @davis2017 findings align with those of @hongslo2009 , who documented a slow regeneration of vegetation cover and species richness in formerly cultivated or grazed communal lands like Concordia. While some areas remained compositionally degraded, others showed recovery particularly where active land management (e.g., fencing, reduced ploughing) had been implemented. This environmental history highlights that observed vegetation trends may not only reflect degradation but also slow ecological recovery due to altered management or climatic shifts @hongslo2009 @davis2017 which could explain the drivers of the notable correlative increase and decline secondary-natural and cropland extents (Figure 18).

```{r}
# Load libraries
library(readxl)
library(dplyr)
library(ggplot2)

# Define file path
file_path <- "C:/Users/K.Mogajane/OneDrive - sanbi.org.za/Kagiso Mogajane/Research Assistant 2025/LandCoverChangeAnalysis_NBA2025/R_Project/nba_landcover/data/Biome_Summaries/Succulent_Karoo.xlsx"

# Read data
SK <- read_excel(file_path)

# Filter data
SK_filtered <- SK %>%
  filter(YEAR %in% c(1990, 2014, 2018, 2020, 2022),
             LC7_text != "Natural"
  )

# Calculate proportional extent by year
SK_filtered <- SK_filtered %>%
  group_by(YEAR) %>%
  mutate(proportion = km2_total / sum(km2_total, na.rm = TRUE)) %>%
  ungroup()

# Define custom colours
custom_colors <- c(
  "Built-up" = "black",
  "Mine" = "red",
  "Artificial-waterbody" = "blue",
  "Secondary-natural" = "orange",
  "Plantation" = "purple",
  "Cropland" = "brown"
)

# Static plot (ggplot2) with line plot and no gridlines
ggplot(SK_filtered, aes(x = YEAR, y = proportion, color = LC7_text, group = LC7_text)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_color_manual(values = custom_colors) +
  labs(
    x = "Year",
    y = "Proportional extent",
    color = "Land cover class"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```

Figure 18. The proportional extent of pressures within the Succulent Karoo biome between 1990 and 2022 land cover sets

```{r}
library(readxl)
library(dplyr)
library(ggplot2)

# -------------------------------
# 1. Read data
# -------------------------------
SK2 <- read_excel("C:/Users/K.Mogajane/OneDrive - sanbi.org.za/Kagiso Mogajane/Research Assistant 2025/LandCoverChangeAnalysis_NBA2025/R_Project/nba_landcover/data/Biome_Summaries/Succulent_Karoo.xlsx")

# -------------------------------
# 2. Filter for 2022 and prepare groups

SK2_2022 <- SK2 %>%
  filter(YEAR == 2022) %>%
  mutate(
    group = ifelse(LC7_text == "Natural", "Natural", "Non-natural"),
    subgroup = ifelse(LC7_text == "Natural", "Natural", LC7_text)
  ) %>%
  group_by(group, subgroup) %>%
  summarise(km2 = sum(km2_total, na.rm = TRUE), .groups = "drop")

# -------------------------------
# 3. Custom colour scheme (hex values)
# -------------------------------
custom_colors <- c(
  "Natural" = "#2E6F40",              # green
  "Built-up" = "#000000",             # black
  "Mine" = "#FF0000",                  # red
  "Artificial-waterbody" = "#0000FF",  # blue
  "Secondary-natural" = "#FFA500",     # orange
  "Plantation" = "#800080",           # purple
  "Cropland" = "#A52A2A"              # brown
)

# -------------------------------
# 4. Create stacked bar plot
# -------------------------------
ggplot(SK2_2022, aes(x = group, y = km2, fill = subgroup)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = custom_colors) +
  labs(
    x = "",
    y = "Area (km²)",
    fill = "Land cover type"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    text = element_text(size = 12)
  )
```

Figure 19. The extent of natural and non natural cover across the biome shows that roughly 80 000 km² of its area remains natural, while less than 20 000 km² has been transformed through various land uses. There is more transformed areas than there are natural areas within the biome. These land use classes are simplified into 7 classes and including the “secondary natural” class for any pixel that transitioned from a non natural class to the natural class. This transformed portion is dominated by cropland, followed by secondary-natural and then mine and built-up areas. While artificial waterbodies and plantations each account for only a small fraction of the transformed landscape.

### Savanna

The Savanna biome is the largest biome in South Africa and covers the majority of sub-Saharan Africa. South African Savannas span a wide rainfall gradient from the arid Kalahari savannas in the northwest to the mesic Zululand savannas in the east, which form part of the Maputaland-Polondoland-Albany biodiversity hotspot @rutherford2006b.

The Savanna biome remains largely intact, with a natural habitat loss between 1990 and 2022 for the biome is 4.1% (Table 1). This loss is primarily driven by the expansion of croplands, which cover a greater extent of the biome than any other land cover pressure (Figure 20). Built-up areas exert the second greatest pressure and continue to expand, while plantations, though showing a slight decline in overall extent since the 2020 land cover, remain a notable driver of habitat transformation within the biome.

From 1990 to 2014, both cropland and plantation areas decreased, while built-up areas and secondary-natural vegetation increased, particularly secondary-natural vegetation. Mining and artificial waterbodies showed slight declines during this period (Figure 20). Between 2014 and 2022, cropland expanded overall, and plantations and artificial waterbodies also increased slightly. Secondary-natural vegetation showed some fluctuation but remained relatively high, while mining declined slightly. Built-up areas remained generally stable (Figure 20).

Across savanna ecosystems, the land cover changes reveal a strong correlative relationship between cropland, secondary natural vegetation, and built-up areas, where increases in cropland often coincide with rising secondary natural and built-up covers, and vice versa. Land use and cover studies on critical biodiversity areas @musetsho2021 dominated by thickets, woodlands, grasslands, and indigenous forests showed a 38.79% decline in thickets and dense bush alongside a 9.43% drop in subsistence agriculture, while woodland cover increased by 46.94% and built-up areas expanded modestly by 1.09%. These shifts reflect key drivers such deforestation, agricultural activities, and urban encroachment that align with broader patterns across the biome which are contributor to the habitat loss @musetsho2021.

The role of agriculture and grazing in shaping savanna vegetation structure and function further explains these land cover patterns. Plant functional responses across a disturbance gradient from maize fields to adjacent rangelands, reveal that intensive farming favours disturbance-adapted ruderal species, while functional diversity remains higher in less disturbed rangelands due to a richer species pool @botha2017. This shift toward opportunistic traits helps explain the increasing secondary natural vegetation class, reflecting that the ecosystems are altered but not entirely lost.

Long-term agricultural transformation creates persistent vegetation states difficult to reverse to original ecological state, driven by feedbacks from altered soil and physical conditions @rutherford2010. Mining and human settlements expanded in other areas within the biome due to demand for mine workers @marais2018, contributing to the growing area of built-up area captured by the land cover.

```{r}
# Load libraries
library(readxl)
library(dplyr)
library(ggplot2)

# Define file path
file_path <- "C:/Users/K.Mogajane/OneDrive - sanbi.org.za/Kagiso Mogajane/Research Assistant 2025/LandCoverChangeAnalysis_NBA2025/R_Project/nba_landcover/data/Biome_Summaries/Savanna.xlsx"

# Read data
SV <- read_excel(file_path)

# Filter data
SV_filtered <- SV %>%
  filter(YEAR %in% c(1990, 2014, 2018, 2020, 2022),
             LC7_text != "Natural"
  )

# Calculate proportional extent by year
SV_filtered <- SV_filtered %>%
  group_by(YEAR) %>%
  mutate(proportion = km2_total / sum(km2_total, na.rm = TRUE)) %>%
  ungroup()

# Define custom colours
custom_colors <- c(
  "Built-up" = "black",
  "Mine" = "red",
  "Artificial-waterbody" = "blue",
  "Secondary-natural" = "orange",
  "Plantation" = "purple",
  "Cropland" = "brown"
)

# Static plot (ggplot2) with line plot and no gridlines
ggplot(SV_filtered, aes(x = YEAR, y = proportion, color = LC7_text, group = LC7_text)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_color_manual(values = custom_colors) +
  labs(
    x = "Year",
    y = "Proportional extent",
    color = "Land cover class"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```

Figure 20. The proportional extent of pressures within the Savanna biome between 1990 and 2022 land cover sets.

```{r}
library(readxl)
library(dplyr)
library(ggplot2)

# -------------------------------
# 1. Read data
# -------------------------------
SV2 <- read_excel("C:/Users/K.Mogajane/OneDrive - sanbi.org.za/Kagiso Mogajane/Research Assistant 2025/LandCoverChangeAnalysis_NBA2025/R_Project/nba_landcover/data/Biome_Summaries/Savanna.xlsx")

# -------------------------------
# 2. Filter for 2022 and prepare groups

SV2_2022 <- SV2 %>%
  filter(YEAR == 2022) %>%
  mutate(
    group = ifelse(LC7_text == "Natural", "Natural", "Non-natural"),
    subgroup = ifelse(LC7_text == "Natural", "Natural", LC7_text)
  ) %>%
  group_by(group, subgroup) %>%
  summarise(km2 = sum(km2_total, na.rm = TRUE), .groups = "drop")

# -------------------------------
# 3. Custom colour scheme (hex values)
# -------------------------------
custom_colors <- c(
  "Natural" = "#2E6F40",              # green
  "Built-up" = "#000000",             # black
  "Mine" = "#FF0000",                  # red
  "Artificial-waterbody" = "#0000FF",  # blue
  "Secondary-natural" = "#FFA500",     # orange
  "Plantation" = "#800080",           # purple
  "Cropland" = "#A52A2A"              # brown
)

# -------------------------------
# 4. Create stacked bar plot
# -------------------------------
options (scipen = 999)
ggplot(SV2_2022, aes(x = group, y = km2, fill = subgroup)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = custom_colors) +
  labs(
    x = "",
    y = "Area (km²)",
    fill = "Land cover type"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    text = element_text(size = 12)
  )
```

Figure 21. The extent of natural and non natural cover across the biome shows that roughly 80 000 km² of its area remains natural, while less than 20 000 km² has been transformed through various land uses. There is more transformed areas than there are natural areas within the biome. These land use classes are simplified into 7 classes and including the “secondary natural” class for any pixel that transitioned from a non natural class to the natural class. This transformed portion is dominated by cropland, followed by secondary-natural and then mine and built-up areas. While artificial waterbodies and plantations each account for only a small fraction of the transformed landscape.

### Azonal Vegetation

The Azonal Vegetation is a fragmented habitat across the country that is shaped by water and salinity. It is found in wetlands, floodplains, salt pans and riparian areas and each ecosystem differs sharply in vegetation. This vegetation plays pivotal ecological role in supporting both widespread species and local endemics especially in biodiversity-rich zones like the Fynbos and Drakensburg @mucina2006e.

This vegetation type is under various pressures because of its wide distribution and its presence across different biomes. Despite this, it remains largely intact, with a recent loss of around 4% (see Table 1). The main pressure on this vegetation, as indicated by land cover data, comes from croplands. Built-up areas also pose a growing pressure on the vegetation.

Between 1990 and 2014, areas under cropland, built-up land, plantations, and mining decreased, while secondary natural vegetation and artificial waterbodies expanded (Figure 22). From 2014 to 2022, cropland and secondary natural areas fluctuated but showed an overall decline, whereas plantations and artificial waterbodies increased slightly. Built-up areas also decreased, and mining remained fairly stable with a slight drop towards the end of the period (Figure 22).

The dominant land use pressure on Azonal Vegetation comes from croplands (Figure 22). Despite a notable reduction in cropland extent in recent years, as seen between 2014 and 2019 land cover data sets (Figure 21), its cumulative impacts remain significant @richardson2007. Agricultural activities adjacent to riparian areas can lead to habitat fragmentation, disrupt natural flooding cycles and increase the likelihood of invasive plant encroachment @richardson2007.

Recurrent droughts exacerbate these agricultural impacts, as they are a significant limiting factor for riparian plants, especially in arid and semi-arid areas where prolonged drought or flow reductions can lower water tables @richardson2007. A study of the Olifants River over a 25-year period by @mashaba2025 illustrates these combined effects, showing how land use and altered water flows drive significant vegetation changes. The research found that heavy livestock grazing led to an increase in grazing-resistant grasses like Couch grass (*Cynodon dactylon)* while palatable grasses such as Red grass (*Themeda triandra)* declined @mashaba2025. Anthropogenic disturbances like agriculture and variable water flows create ideal conditions for invasive species to establish and proliferate @richardson2007; @mashaba2025. This was confirmed in the Olifants River study, which documented the spatial increase of invasive alien forbs Tall verbena (*Verbena bonariensis)* and woody plants Bloubasboom and Brasilean Glory Pea (*Acacia dealbata* and *Sesbania punicea,* respectively) over the 25-year period, often linked to disturbances providing suitable substrate @mashaba2025. Both floods and droughts impact these ecosystems; floods remove vegetation and deposit sediment, creating new habitats that can facilitate the incursion of alien plants @richardson2007; @mashaba2025

South Africa’s timber plantations are located either within or next to indigenous forest and the Indian Ocean Coastal Belt @karumbidza2006 and sometimes Azonal Vegetation sits near the Forest ecosystems and as a result are impacted by introduction of plantation. For example, in KwaZulu-Natal, in the Richard’s Bay area, there are industrial timber plantation that are where Azonal Vegetation has been mapped in the National Vegetation Map @mucina2006 @dayaram2019. These plantation have led to habitat loss due to conversion of natural land to monoculture plantations, which has associated impacts in reduced streamflow and impacting on indigenous aquatic species @karumbidza2006. Although at the time of publication @karumbidza2006 there was very little proportion of plantation in the Azonal Vegetation (Figure 22); there increase in secondary-natural extent may suggest that the plantations.

```{r}
# Load libraries
library(readxl)
library(dplyr)
library(ggplot2)

# Define file path
file_path <- "C:/Users/K.Mogajane/OneDrive - sanbi.org.za/Kagiso Mogajane/Research Assistant 2025/LandCoverChangeAnalysis_NBA2025/R_Project/nba_landcover/data/Biome_Summaries/Azonal.xlsx"

# Read data
AZ <- read_excel(file_path)

# Filter data
AZ_filtered <- AZ %>%
  filter(YEAR %in% c(1990, 2014, 2018, 2020, 2022),
             LC7_text != "Natural"
  )

# Calculate proportional extent by year
AZ_filtered <- AZ_filtered %>%
  group_by(YEAR) %>%
  mutate(proportion = km2_total / sum(km2_total, na.rm = TRUE)) %>%
  ungroup()

# Define custom colours
custom_colors <- c(
  "Built-up" = "black",
  "Mine" = "red",
  "Artificial-waterbody" = "blue",
  "Secondary-natural" = "orange",
  "Plantation" = "purple",
  "Cropland" = "brown"
)

# Static plot (ggplot2) with line plot and no gridlines
ggplot(AZ_filtered, aes(x = YEAR, y = proportion, color = LC7_text, group = LC7_text)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_color_manual(values = custom_colors) +
  labs(
    x = "Year",
    y = "Proportional extent",
    color = "Land cover class"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```

Figure 22. The proportional extent of pressures within the Azonal Vegetation biome between 1990 and 2022 land cover sets.

```{r}
library(readxl)
library(dplyr)
library(ggplot2)

# -------------------------------
# 1. Read data
# -------------------------------
AZ2 <- read_excel("C:/Users/K.Mogajane/OneDrive - sanbi.org.za/Kagiso Mogajane/Research Assistant 2025/LandCoverChangeAnalysis_NBA2025/R_Project/nba_landcover/data/Biome_Summaries/Azonal.xlsx")

# -------------------------------
# 2. Filter for 2022 and prepare groups

AZ2_2022 <- AZ2 %>%
  filter(YEAR == 2022) %>%
  mutate(
    group = ifelse(LC7_text == "Natural", "Natural", "Non-natural"),
    subgroup = ifelse(LC7_text == "Natural", "Natural", LC7_text)
  ) %>%
  group_by(group, subgroup) %>%
  summarise(km2 = sum(km2_total, na.rm = TRUE), .groups = "drop")

# -------------------------------
# 3. Custom colour scheme (hex values)
# -------------------------------
custom_colors <- c(
  "Natural" = "#2E6F40",              # green
  "Built-up" = "#000000",             # black
  "Mine" = "#FF0000",                  # red
  "Artificial-waterbody" = "#0000FF",  # blue
  "Secondary-natural" = "#FFA500",     # orange
  "Plantation" = "#800080",           # purple
  "Cropland" = "#A52A2A"              # brown
)

# -------------------------------
# 4. Create stacked bar plot
# -------------------------------
ggplot(AZ2_2022, aes(x = group, y = km2, fill = subgroup)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = custom_colors) +
  labs(
    x = "",
    y = "Area (km²)",
    fill = "Land cover type"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    text = element_text(size = 12)
  )
```

Figure 23. The extent of natural and non natural cover across the vegetation shows that roughly 20 000 km² of its area remains natural, while less than 5 000 km² has been transformed through various land uses. There is more transformed areas than there are natural areas within the biome. These land use classes are simplified into 7 classes and including the “secondary natural” class for any pixel that transitioned from a non natural class to the natural class. This transformed portion is dominated by cropland, followed by secondary-natural and then artificial waterbodies and built-up areas. While mining and plantations each account for only a small fraction of the transformed landscape.
:::

## Approach

This report summarises trends in the extent of land cover change for every biome from 1990 to 2022 Land Use and Land Cover data sets. The aim of this report is to track loss of natural habitat and the drivers behind the loss.

We use the National Land Cover 7 class version prepared by the South African National Biodiversity Institute (SANBI) for 1990, 2014, 2018, 2020 ad 2022, based on the National Land Cover products served by the Department of Forestry, Fisheries and the Environment), reclassified by SANBI from 72 classes to 7 classes. The re-classified National Land Cover 7 classes are described as follows: 1 = Natural or near natural ecosystems

2 = Secondary-natural (previously cropland or mine or built-up, now re-vegetated partially)

3 = Artificial-waterbodies (dams and water treatment)

4 = Built-up (villages, urban, industrial, commercial and infrastructure)

5 = Cropland

6 = Mine

7 = Plantation (alien timber and pulp plantations)

Below we summaries description of each land cover class used in this report and whether they are classified as natural or non-natural.

| Land cover class | Description | Natural/ Non-natural |
|------------------------|------------------------|------------------------|
| Natural | Areas still in a natural or near-natural state, including indigenous grasslands, shrublands, forests, woodlands, natural bare areas, and natural water bodies | Natural |
| Secondary-natural | Previously transformed land (e.g., abandoned croplands or mines) now undergoing passive recovery. Vegetation may be returning, but the land remains altered. | Non-natural |
| Artificial-waterbodies | Man-made water features such as dams, canals, sewage ponds, and flooded mine pits, distinct from natural lakes or rivers. | Non-natural |
| Croplands | Land actively used for crop production, including field crops, orchards, vineyards, pivot and non-pivot irrigation, and intensive farming such as feedlots. | Non-natural |
| Built-up | Includes all settlements (formal, informal, villages), commercial, and industrial areas, with roads, buildings, and other infrastructure replacing natural cover. | Non-natural |
| Plantation | Areas planted with non-native tree species for timber production, including mature plantations and recently harvested stands. | Non-natural |

: Table 2. Descriptions of the land cover classes used in this report and whether the class is classified as natural or non-natural.
